Efficient pedigree recording for fast population genetics simulation

### 2017-12-09 forwards_paper.tex
a few more edits

To measure the performance gains from recording the pedigree we ran simulations with and
without recording, for both implementations (\simupop{} \citep{simupop}
and \fwdpp{} \citep{fwdpp}, as described below).
All simulations implemented a discrete-time Wright-Fisher population of diploid
population size $N$, simulated for $10N$ generations.
The simulations without pedigree recording introduced neutral mutations at a rate 
equal to the recombination rate,
so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations and recombinations,
respectively, per gamete and per generation.
% Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
% and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
We ran simulations with different values of $N$
and of the \emph{scaled} parameter $\rho = 4Nr$,
categorized by the value of $\rho$, since this
is what most strongly determines run time
(along with the scaled mutation rate, $\theta = 4N\mu$, which is also equal to $\rho$.)

@@

@@

\subsubsection*{The \fwdpp{} implementation}

We implemented Wright-Fisher simulations in C++ using \fwdpp{} library
functions and interface code that records data into the \msprime{} Tables API (see below).
In these simulations, $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
\citep{Kimura1969-uc}.
We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.

\plr{Perhaps implementations should go below in Methods?}
\plr{Insert figures/tables of performance here and describe them.}

\subsubsection*{The \simupop{} implementation}

We implemented Wright-Fisher simulations in Python using \simupop{} library
functions (which wrap C++ code) and interface code (\ftprime{}, also in Python) that records data into the \msprime{} Tables API (see below).
In these simulations, $\mu$ and $r$ define a per-base-pair mutation rate and we
defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The left-hand panel shows the tree sequence pictorially in two different ways:
        (top) as a sequence of tree topologies
        and (bottom) the spatial extent of the edges that define these topologies.
        The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \plr{Whoops!  Genotypes in the second tree should be `GGC'}
        \jk{Some minor alignment issues present here.}
        \plr{turns out this is no fun to edit in inkscape.
            I can re-make to match the other figs
            but will leave alone for now: it's good!}
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/num_edges_50}
        \includegraphics{sims/simplify_timings}
    \end{center}
    \caption{
        \textbf{(A)}
        Number of edges in the simplified tree sequence
        for 50 Wright--Fisher simulations begun with no prior history,
        as a function of number of generations.
        Each line is one simulation, the black line gives the average,
        and the dotted line is the upper bound of equation \eqref{eqn:edge_bound}.
        \plr{Propose to add other figures as subfigures here:}
        \textbf{(B)}
        Time required to simplify the first $k$ edges
        of \plr{what tree sequence?},
        plotted against $k$.
        \textbf{(C)}
        Time required to simplify
        \plr{what tree sequence?}
        down to a sample of size $n$,
        plotted against $n$
        (note the log scale;
        the time scales logarithmically with $n$).
        \label{fig:simplify_complexity}
    }
\end{figure}

@@

\plr{Replace this with Jerome's implementation.}

>>>>

To measure the performance gains from recording the pedigree we ran simulations both with and without recording.
All simulations implemented a discrete-time Wright-Fisher population of diploid
population size $N$, simulated for $10N$ generations.
The simulations without pedigree recording introduced neutral mutations at a rate 
equal to the recombination rate,
so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations and recombinations,
respectively, per gamete and per generation.
% Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
% and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
We ran simulations with different values of $N$
and of the \emph{scaled} parameter $\rho = 4Nr$,
categorized by the value of $\rho$, since this
is what most strongly determines run time
(along with the scaled mutation rate, $\theta = 4N\mu$, which is also equal to $\rho$.)

@@

To demonstrate the method's flexibility and to test it in different contexts,
we implemented these simulations separately using two different simulation engines:
\begin{enumerate}

    \item 
        In C++, using \fwdpp{} library functions and interface code (details below),
        using a continuum-sites model for both mutation and recombination.
        % $\mu$ and $r$ were the means of Poisson processes,
        % and mutations used the infinite-sites model \citep{Kimura1969-uc}.
        % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.

    \item 
        In Python, using the \simupop{} library interfacing with a custom Python library (\ftprime{}),
        using a discrete but large number of sites.
        % $\mu$ and $r$ define per-base-pair mutation rate and we
        % defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
        % We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.

\end{enumerate}

@@

\plr{Insert figures/tables of performance here and describe them.}

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The left-hand panel shows the tree sequence pictorially in two different ways:
        (top) as a sequence of tree topologies
        and (bottom) the spatial extent of the edges that define these topologies.
        The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/num_edges_50}
        \includegraphics{sims/simplify_timings}
    \end{center}
    \caption{
        \textbf{(A)}
        Number of edges in the simplified tree sequence
        for 50 Wright--Fisher simulations begun with no prior history,
        as a function of number of generations.
        Each line is one simulation, the black line gives the average,
        and the dotted line is the upper bound of equation \eqref{eqn:edge_bound}.
        \plr{Propose to add other figures as subfigures here; but better than this.}
        \textbf{(B)}
        Time required to simplify the first $k$ edges
        of \plr{what tree sequence?},
        plotted against $k$.
        \textbf{(C)}
        Time required to simplify
        \plr{what tree sequence?}
        down to a sample of size $n$,
        plotted against $n$
        (note the log scale;
        the time scales logarithmically with $n$).
        \label{fig:simplify_complexity}
    }
\end{figure}

@@

\plr{Replace this with Jerome's interval-tree implementation?}
